---
title: "Assignment 1 - GSE263125 Analysis"
output: html_document
date: "2025-02-09"
---

# **Title**: Assignment 1 - GSE263125 Analysis
# **Name**: Soomi Choi

# **1. Introduction**

This is notebook analyzes the dataset **GSE263125** and investigates the impact of gut inflammation on enteric neurodegeneration. The dataset includes both human and mouse samples that represent control and test conditions, but we're only looking at human samples here.

# **2. Data Source**

- **Accession Number**: GSE263125
- **Organism**: Homo sapiens
- **Publication Date**: Dec 31, 2024
- **Number of Human Samples**: 8

# **3. Methods**

## **3-1. Load the Libraries**

```{r setup, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("GEOquery", "limma", "org.Hs.eg.db", "AnnotationDbi", "edgeR"))
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages(c("tidyverse", "rmarkdown"))

library(GEOquery)
library(limma)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(tidyverse)
library(edgeR)
library(rmarkdown)
```

## **3-2. Download the Data**

```{r download}
gse_id <- "GSE263125"
data_dir <- "data"

if (!dir.exists(data_dir)) dir.create(data_dir)

gse_file <- file.path(data_dir, paste0(gse_id, ".rds"))

if (!file.exists(gse_file)) {
  gse <- getGEO(gse_id, GSEMatrix = TRUE, getGPL = TRUE)
  saveRDS(gse, gse_file)
} else {
  gse <- readRDS(gse_file)
}

gse <- gse[["GSE263125-GPL24676_series_matrix.txt.gz"]]
metadata <- pData(gse)
knitr::kable(head(metadata), caption = "Metadata from GEO")
```

This selection below downloads, extracts, and loads the supplementary count matrix from GEO, ensuring the expression data is available for preprocessing and analysis.

```{r install}
library(GEOquery)

# Define data directory
data_dir <- "data"
if (!dir.exists(data_dir)) dir.create(data_dir)

# Download the supplementary file
supplementary_url <- "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE263nnn/GSE263125/suppl/GSE263125_Chat-Ribo_counts.txt.gz"
supp_file <- file.path(data_dir, "GSE263125_Chat-Ribo_counts.txt.gz")

if (!file.exists(supp_file)) {
    download.file(supplementary_url, destfile = supp_file, mode = "wb")
}

# Unzip the file
unzipped_file <- gsub(".gz", "", supp_file)
if (!file.exists(unzipped_file)) {
    print("🔹 Extracting count matrix...")
    R.utils::gunzip(supp_file, remove = FALSE)
}

# Load count matrix
expr_data <- read.delim(unzipped_file, row.names = 1)
```

## **3-3. Filter Human Samples & Map Genes to HUGO Symbols**

```{r filter}
library(stringdist)

# 1. Standardize sample names in metadata and expression data
metadata$geo_accession <- trimws(toupper(metadata$geo_accession))
metadata$description <- trimws(toupper(metadata$description))
colnames(expr_data) <- trimws(toupper(colnames(expr_data)))

# 2. Identify human samples
human_samples <- metadata$geo_accession[metadata$taxid_ch1 == "9606"]
if (length(human_samples) == 0) stop("No human samples found")

# 3. Map GEO Accession to Sample Name
human_sample_names <- na.omit(setNames(metadata$description, metadata$geo_accession)[human_samples])
if (length(human_sample_names) == 0) stop("o valid human sample names mapped")

# 4. Match sample names to expr_data
valid_samples <- intersect(human_sample_names, colnames(expr_data))

# 5. Attempt fuzzy matching if no direct matches
if (length(valid_samples) == 0) {
    dist_matrix <- stringdistmatrix(human_sample_names, colnames(expr_data), method = "jw")  
    best_matches <- apply(dist_matrix, 1, function(row) colnames(expr_data)[which.min(row)])
    
    best_matches_df <- data.frame(Human_Sample = human_sample_names, 
                                  Matched_Column = best_matches, 
                                  Distance = apply(dist_matrix, 1, min))

    best_matches_df <- best_matches_df[best_matches_df$Distance <= 0.15, ]
    best_matches_df <- best_matches_df[!duplicated(best_matches_df$Matched_Column), ]
    
    valid_samples <- best_matches_df$Matched_Column
}

# 6. Filter expression data
if (length(valid_samples) > 0) {
    expr_data <- expr_data[, valid_samples, drop = FALSE]
} else {
    stop("o valid human samples found in expr_data")
}
```

## **3-4. Assess Data Quality**

```{r assess}
# 1. Check if expr_data is empty
if (ncol(expr_data) == 0 | nrow(expr_data) == 0) {
    stop("expr_data is empty")
}

# 2. Compute Summary Statistics
summary_stats <- data.frame(
  Sample = colnames(expr_data),
  Mean = colMeans(expr_data, na.rm = TRUE),
  Median = apply(expr_data, 2, median, na.rm = TRUE),
  SD = apply(expr_data, 2, sd, na.rm = TRUE)
)

# 3. Display Summary Statistics Table
knitr::kable(summary_stats, caption = "Summary Statistics of Expression Data")

# 4. Boxplot for Data Distribution
boxplot(expr_data, 
        main = "Distribution of Expression Data", 
        las = 2, 
        col = "lightpink", 
        xlab = "Samples", 
        ylab = "Expression Value", 
        outline = FALSE)

# 5. Identify and remove all-zero samples
expr_data <- expr_data[, colSums(expr_data) > 0, drop = FALSE]
```

## **3-5. Clean the Data (Remove Outliers)**

```{r clean}
# 1. Check if `expr_data` is empty before proceeding
if (ncol(expr_data) == 0 | nrow(expr_data) == 0) {
    stop("`expr_data` is empty before outlier removal")
}

# 2. Compute MAD scores (Median Absolute Deviation for each sample)
mad_scores <- apply(expr_data, 2, mad, na.rm = TRUE)

# 3. Define dynamic outlier threshold (adaptive)
outlier_threshold <- 3 * median(mad_scores, na.rm = TRUE)

# 4. Identify outlier samples
outlier_samples <- which(mad_scores > outlier_threshold)

# 5. Ensure `expr_data` does not become empty after removing outliers
if (length(outlier_samples) > 0 && length(outlier_samples) < ncol(expr_data) - 1) {
    expr_data <- expr_data[, -outlier_samples, drop = FALSE]
    print(paste0("Removed ", length(outlier_samples), " outlier samples."))
} else {
    print(" Too many outliers detected")
}

# 6. Ensure metadata remains aligned
metadata <- metadata[metadata$geo_accession %in% colnames(expr_data), , drop = FALSE]

# 7. Ensure `expr_data` is not empty before visualization
if (ncol(expr_data) == 0 | nrow(expr_data) == 0) {
    stop("expr_data is empty after outlier removal")
}

# 8. Convert `expr_data` to numeric if necessary
if (!all(sapply(expr_data, is.numeric))) {
    expr_data <- as.data.frame(lapply(expr_data, as.numeric))
    print("Converted `expr_data` to numeric format.")
}

# 9. Boxplot After Cleaning
boxplot(expr_data, 
        main = "Distribution of Expression Data After Cleaning", 
        las = 2, 
        col = "lightgreen", 
        xlab = "Samples", 
        ylab = "Expression Value", 
        outline = FALSE)
```

## **3-6. Map Probes to Gene Symbols**

```{r map_genes}
# 1. Load feature data
feature_data <- featureData(gse)@data

# 1-1. If feature data is empty, fetch from platform annotation
if (ncol(feature_data) == 0) {
    platform_id <- "GPL24676"
    gpl <- getGEO(platform_id)
    feature_data <- Table(gpl)
}

# 2. Determine organism
organism <- ifelse(any(grepl("RIK$", rownames(expr_data))), "mouse", "human")

# 3. Load appropriate annotation database & define keytype_guess
if (organism == "mouse") {
    if (!requireNamespace("org.Mm.eg.db", quietly = TRUE)) {
        BiocManager::install("org.Mm.eg.db")
    }
    library(org.Mm.eg.db)
    db <- org.Mm.eg.db
    keytype_guess <- "SYMBOL"
} else {
    if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
        BiocManager::install("org.Hs.eg.db")
    }
    library(org.Hs.eg.db)
    db <- org.Hs.eg.db
    keytype_guess <- "SYMBOL"
}

# 4. Map IDs to gene symbols
p_to_s <- AnnotationDbi::select(
    db, 
    keys = rownames(expr_data), 
    keytype = "SYMBOL", 
    columns = c("SYMBOL", "GENENAME")
)

# 5. Convert to a data frame if needed
p_to_s <- as.data.frame(p_to_s)

# 6. Ensure valid mapping
if (!"SYMBOL" %in% colnames(p_to_s)) {
    stop("No 'SYMBOL' column found in mapping results")
}

# 7. Add ID column manually (if missing)
p_to_s$ID <- rownames(expr_data)

# 8. Remove missing and duplicated values
p_to_s <- na.omit(p_to_s)
p_to_s <- p_to_s[!duplicated(p_to_s$SYMBOL), ]

# 9. Match row names of expr_data with mapped IDs
expr_data <- expr_data[rownames(expr_data) %in% p_to_s$ID, , drop = FALSE]
rownames(expr_data) <- p_to_s$SYMBOL[match(rownames(expr_data), p_to_s$ID)]

# 10. Remove duplicated gene symbols
expr_data <- expr_data[!duplicated(rownames(expr_data)), ]
```

# **4. Normalization**

```{r normalize}
# 1. Apply Log2 transformation with +1 offset
expr_data <- log2(expr_data + 1)

# 2. Boxplot visualization of normalized data
boxplot(expr_data, 
        main = "Expression Data After Normalization", 
        las = 2, 
        col = "lightblue", 
        xlab = "Samples", 
        ylab = "Log2 Expression Value", 
        outline = FALSE)
```

# **5. Final Result**

```{r final-result}
# 1. Ensure row names are correctly mapped to gene symbols
matched_genes <- p_to_s$SYMBOL[match(rownames(expr_data), p_to_s$ID)]
valid_genes <- !is.na(matched_genes)

# 2. Keep only valid mappings
expr_data <- expr_data[valid_genes, , drop = FALSE]
rownames(expr_data) <- matched_genes[valid_genes]

# 3. Remove duplicated gene symbols
expr_data <- expr_data[!duplicated(rownames(expr_data)), ]

# 4. Define output directory and save processed data
output_dir <- "data/processed"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(expr_data, file.path(output_dir, "GSE263125_processed.csv"), row.names = TRUE)
```

